---
title: "For submission to the Journal of Integrated Pest Management"
subtitle: "Category: 'Issues'"
output: 
  beamer_presentation: 
    latex_engine: xelatex
bibliography: 'thesis.bib'
csl: 'j_med_entomol.csl'
#always_allow_html: true
---
```{r setup, include=FALSE}
#setup options
knitr::opts_chunk$set(fig.width=12, fig.height=8, echo=FALSE)

#a list of packages used for scripts
pkgs <- c('captioner', 'pryr', 'tiff')

#loading required packages
lapply(pkgs, library, character.only = TRUE)
rm(pkgs)

#setting up captions for use in the document
fig_nums <- captioner::captioner(prefix = 'Fig.')
tab_nums <- captioner::captioner(prefix = 'Table')

#allows me to use a custom function instead of the citation function of captioner, which is wordy to use
citef <- pryr::partial(fig_nums, display = "cite")
citet <- pryr::partial(tab_nums, display = "cite")
```

### First report of the *Brevipalpus*-transmitted (Trombidiformes: Tenuipalpidae) virus, *Orchid fleck virus* (Mononegavirales: Rhabdoviridae) infecting three ornamentals in Florida

Austin **Fife**^1^, Daniel **Carrillo**^2^, Gary **Knox**^3^, Fanny **Iriarte**^4^, Kishore **Dey**^5^, Avijit **Roy**^6^, Ronald **Ochoa**^7^, Gary **Bauchan**^8^ $\dagger$, Mathews **Paret**^4^, Xavier **Martini**^1^\*

^1^ University of Florida, Department of Entomology and Nematology, North Florida Research and Education Center, Quincy FL 32351

^2^ University of Florida, Department of Entomology and Nematology, Tropical Research and Education Center, Homestead FL 33031

^3^ University of Florida, Department of Environmental Horticulture, North Florida Research and Education Center, Quincy FL 32351

^4^ University of Florida, Department of Plant Pathology, North Florida Research and Education Center, Quincy FL 32351

^5^ The Florida Department of Agriculture and Consumer Services, Division of Plant Industry, Section of Plant Pathology, Doyle Conner Building, 1911 SW 34th street, Gainesville, FL 32608

^6^	United States Department of Agriculture – Agriculture Research Service, Molecular Plant Pathology Laboratory, 10300 Baltimore Ave, Bldg. 4 BARC-West, Beltsville, MD 20705

^7^ United States Department of Agriculture - Agriculture Research Service, Systematic Entomology Laboratory 10300 Baltimore Ave, Bldg. 5 BARC-West, Beltsville, MD 20705

^8^	United States Department of Agriculture - Animal and Plant Health Inspection Service, Electron and Confocal Microscopy Unit, Bldg. 12  BARC-West, 10300 Baltimore Ave, Beltsville, MD 20705

*Corresponding author; E-mail: xmartini\@ufl.edu Phone: 850-875-7160 Fax: 352-846-6617

\pagebreak

### Abstract
Several flat mite species, all from the genus *Brevipalpus* Donnadieu (Trombidiformes: Tenuipalpidae), are the only known vectors of dichorhaviruses. The *B. californicus* species group exclusively can transmit Orchid fleck dichorhavirus (OFV) in a persistent propagative manner. OFV is the type species for the genus *Dichorhavirus* and infects more than fifty plant species belongs to the family Orchidaceae, Asparagaceae (Nolinoidaea), and Rutaceae (*Citrus*). During June 2020, chlorotic ringspot symptoms on Giant Lilyturf (*Liriope* spp., cv. ‘Gigantea’) were observed in a landscape in Leon County, Florida. Later in the year, the presence of OFV was confirmed using OFV specific conventional reverse transcription polymerase chain assay (RT-PCR) assay and Sanger sequencing. RT-PCR amplicons had a 98\% identity with the known OFV sequences available in the Genbank. The identification was also confirmed with the quantitative RT-PCR (RT-qPCR). Additional samples were taken from other Nolinoidaea, including *Liriope muscari*, *Ophiopogon japonicus*, *O. intermedius* and *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea) in Leon and Alachua counties. Identification of partial genome sequence confirmed the presence of both the orchid strains (OFV-Orc1 and OFV-Orc2) in Florida. Three mite species were recovered from OFV-infected plants: *Brevipalpus californicus* sensu lato, *B. obovatus* Donnadieu and *B. confusus* Banks. One of these species is presumably responsible for OFV transmission. Florida has various mite species of *Brevipalpus* and a diverse array of susceptible native and introduced plant species in the landscape. We suggest that OFV already has a wide distribution in Florida which will continue to spread if unchecked, representing a potential threat for *Liriope* spp., *Ophiopogon* spp. and *Aspidistra elatior* which are commonly used in landscaping in Florida . In this study we are reporting three new hosts from the family Asperagaceae. To know the extended host range of OFVs, a survey in the citrus growing regions in Florida is essential with special emphasis to the plants belongs to the family Rutaceae and Asparagaceae.


### Background
*Orchid fleck virus* (OFV) is the type member for the genus *Dichorhavirus*, family Rhabdoviridae; a bacilliform, nuclear rhabdovirus composed of two segments of single-stranded, negative-sense RNA which infects plants [@Amarasinghe2019; @Dietzgen2014; @Walker2018]. Other members of this genus are: *Citrus chlorotic spot virus*, *Citrus leprosis virus* N, *Clerodendrum chlorotic spot virus* and *Coffee ringspot virus* [@Dietzgen2018a]. Flat mites from the genus *Brevipalpus* Donnadieu (Trombidiformes: Tenuipalpidae) are the only known vector for dichorhaviruses [@Maeda1998], and *Brevipalpus californicus* (Banks) mites are the only mite known to transmit OFV in a persistent propagative manner [@Kondo2003].

### Symptoms
OFV-infected plants exhibit various symptoms dependent on the infected plant species as well as the strain of the OFV [@Kubo2009b], but symptoms typically appear as chlorotic flecks, which ultimately coalesce into larger spots or ringspot patterns (`r citef("fig_1")`).
<!-- These lesions are thought to be associated with the feeding sites of the mite vector [@cite?]. Also, is it possibly a sign of RNAi plant defenses? See Hull Plant Virology -->

### Distribution
OFV was first described infecting *Cymbidium* orchids in Japan [@Doi1977]. Many countries have reported OFV and OFV-like rhabdoviruses infecting orchids worldwide [@Kondo2003], including Asia: [China [@Peng2017], Korea [@Chang1991, @Zheng2013]], Africa: [South Africa [@Blanchfield2001]], North America: [The United States [@Blanchfield2001; @Bratsch2015]], South America: [Brazil [@Kitajima1974; @Kitajima2001], Colombia [@Kubo2009b], Costa Rica [@FreitasAstua2002], Paraguay [@RamosGonzalez2015]], Europe: [Denmark [@Begtrup1972], France [@Sauvetre2018], Germany [@Lesemann1975; @Petzold1971]] and Oceania: [Australia [@Lesemann1971; @Lesemann1975; @Gibbs2000], Fiji [@Pearson1993], Vanuatu [@Pearson1993]]. The prevalence of OFV and its mite vector is thought to be associated with the movement of infected orchids [@Dietzgen2018a].

### Infectivity
OFV naturally infects more than fifty species of Orchidaceae [@Kitajima2010; @Peng2013], some Asparagaceae (Nolinoidaea) [@Mei2016; @Dietzgen2018b], as well as Rutaceae: *Citrus* where it causes citrus leprosis-like symptoms [@Beltran-Beltran2020; @GarciaEscamilla2018; @Roy2020]. Mechanical transmission of OFV is possible under lab conditions to various Chenopodiaceae, Aizoaceae, Fabaceae, and Solanaceae [@Chang1976; @Kondo2003; @Peng2013].

### Detection
During June 2020, chlorotic ringspot symptoms were observed on Giant Lilyturf *Liriope* spp., cv. 'Gigantea' in a landscape of Leon County, Florida (`r citef("fig_1")`). *Liriope* belong to a group of plants in the family Asparagaceae, subfamily Nolinoidaea, which includes a diverse array of various monocotyledonous lilliod plants which are native to southeastern Asia [@Chase2009; @Meng2021]. *Liriope* and the closely related *Ophiopogon* Ker Gawler (Asparagaceae: Nolinoidaea) are considered the most important ground cover sold by the nursery industry in southeastern US [@Mcharo2003].

### Testing
Viral infections of suspected plant leaf samples were initially tested at the Plant Disease Diagnostic Clinic at the North Florida Research and Education Center (NFREC) in Quincy, FL. All the samples tested negative for begomovirus, potyvirus, tospovirus as well as for Impatiens necrotic spot virus, Tobacco mosaic virus and Tomato spotted wilt virus. The infected materials were subsequently sent to the Florida Department of Agriculture and Consumer Services (FDACS). The presence of OFV was confirmed using OFV generic R2-Dicho-GF and R2-Dicho-GR primers [@Roy2020] by one step conventional RT-PCR, amplifying ~800 nt of L-gene (RNA2) amplicon from an infected *Liriope* leaf sample. The RT-PCR amplicon was sequenced using Sanger sequencing and confirmed the presence of OFV in the tested samples. Sanger sequencing of RT-PCR amplicons shared 98\% nucleotide identity with orchid strains of OFV (GenBank Accession numbers: AB244418 and LC222630) [@Kondo2006; @Kondo2017]. Further surveys of putatively OFV-infected plants were taken during subsequent visits to the initial site of collection. Plant samples included various cultivars of Nolinoid plants, including *Liriope* spp., *Ophiopogon* spp., and *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea) which were located at the same site as the original infected *Liriope* spp., cv 'Gigantea'. *A. elatior* infected with OFV have chlorotic leaves or chlorotic flecks (`r citef("fig_1")`). No ringspots have been observed on *A. elatior* to date. All plant samples were tested via RT-PCR at the NFREC<!-- (`r citet("tab_1")`) --> to confirm the presence of OFV.

### Map
Further surveys of plants belonging to the subfamily Nolinoidaea in Florida have revealed more sites with symptomatic plants in both Leon and Alachua counties. In addition to the discovery of the OFV-infected plants, we collected mites from symptomatic plants in both Leon and Alachua counties. In addition to the discovery of the OFV-infected plants, we collected mites from symptomatic plants in Leon county which were observed with phase contrast microscopy. We encountered both an unidentified eriophyoid mite species on the *Liriope* spp., along with flat mites on all species tested (*Liriope* spp., *Ophiopogon* spp., and *A. elatior*). The flat mites were originally identified as *Brevipalpus californicus* (Banks) sensu lato, an identity which was confirmed by the FDACS via Differential Interference Contrast (DIC) microscopy. *Brevipalpus* mites have been previously associated with OFV [@Beltran-Beltran2020; @GarciaEscamilla2018; @Dietzgen2018b] as well as similar diseases [@Kitajima2010] and are known to feed on a large variety of economically important plants [@Childers2003a; @Akyazi2017]. Unfortunately, the *Brevipalpus* mite species complex is known to contain cryptic species [@Childers2011] whose identification can be improved with molecular methods as well as more advanced microscopy techniques, such as cryo scanning electron microscopy (currently known as Cryo-SEM) [@Skoracka2015; @Leon2010; @Beard2015]. With that in mind, we collected additional samples   of the mites from the same site as the original OFV detection and sent them to the USDA-ARS in Beltsville to observe the mites with Cryo-SEM  techniques, which agreed with both prior identifications of *B. californicus* s.l. <!-- (`r citef("fig_2")`) -->but revealed the presence of two other species: *B. obovatus* and *B. confusus*.

### Previous Reports
The first report of OFV in the United State is thought to be @Ko1985, who describes nuclear inclusions caused by an undescribed bacilliform rhabdovirus in *Brassia* orchids.  The significance of this report is their reference to spoke-wheel configurations of the viral particles [@Ko1985], a sign typically associated with OFV infection [@Chang1976]. Unfortunately, @Ko1985 made no mention of mites or further investigations of this virus. The first certain report of OFV in the US was made by @Bratsch2015, who confirmed the presence of OFV in *Phalaenopsis* hybrids in the US, using TEM of ultrathin sections of plant tissue as well as molecular sequence analysis and its association with *Brevipalpus* mites. The authors did not make a conclusive species identification but suggested the mite species was *B. californicus* group as the vector as referred by @Kondo2003.

OFV has been reported in other Nolinoidaea in Australia [@Dietzgen2018b; @Mei2016], including *Liriope spicata* (Thunb.) Lour. [@Mei2016]. The Florida collected plants of *Liriope* spp., cv 'Gigantea' are thought to belong to either *Liriope muscari* or *Liriope gigantea*. In addition, we are not aware of any record other than this manuscript which reports OFV infection in *Ophiopogon* plants. Unfortunately, the Ophiopogonae group includes species of *Liriope* and *Ophiopogon* which are very similar in appearance and growth habit, with few useful morphological characters used for their classification [@Fantz2008a]. Furthermore, the horticultural industry has created a diverse array of cultivars of these plants, which are often mislabeled [@Fantz2008a]. Aside from the taxonomic confusion created by humans, natural hybrids between *Ophiopogon* and *Liriope* have created a natural source of error for reconstructing phylogenies [@Zhou2009]. Together, these factors make it difficult to differentiate and identify plants in the landscape by visual inspection alone. These obfuscations of species identity may be accounted for in the future via sequence comparisons of the OFV-infected plants [@Rudall2000; @Wang2014], but these comparisons are beyond the scope of our current report. Nonetheless, we are confident that ours is the first report of OFV infecting *A. elatior*; although [@Zheng2013] mentions the association of *B. californicus* with *A. elatior* they make no mention of OFV symptoms in this plant.


<!-- ### Control methods for *Brevipalpus* mites -->

<!-- <!-- Reducing the populations of the *Brevipalpus* mite vector may be able to reduce the spread of OFV by infective mites, but infected plants are not able to recover from the virus and must be rouged or removed. Rouging plants is a direct and efficient option to remove mite and virus populations quickly but increases in difficultly with the number of plants involved. Once infected plants have been removed, the objective will be to prevent re-infestation of the *Brevipalpus* vector via conventional pest management methods. -->

<!-- <!-- Most research on the integrated management of mites has been catered towards controlling various species of spider mites (Acari: Tetranychidae) [@Leeuwen2015], but many of those methods can be applied to *Brevipalpus* spp. as well: -->
<!-- Chemical control via synthetic acaricides is commonly used to manage *Brevipalpus* spp. [@Leeuwen2015], but acaricides have some drawbacks. Many products control a broad class of arthropods, but often they do so indiscriminately, killing beneficial insects as well as pests [@Suckling2013]. Control strategies which rely heavily on chemical control have the potential to encourage pesticide resistance in mite populations and pesticide resistance has been reported in various *Brevipalpus* populations [@Alves2000; @Campos2002; @Omoto2000; @Rocha2021]. In addition, it is important to consider the interactions which may occur between different chemical applications or tank mixes [@Vechia2018]. Lastly, application costs of chemical controls can be large and increase the cost of production [@Rodrigues2000a], nevertheless chemical applications are valuable tools used to consider against *Brevipalpus*. -->

<!-- Various biological control methods have also been studied, but their use has not been widely adopted [@Messing2017]. Even so, some studies have found biological control methods which could be used specifically for *Brevipalpus* mites: The predatory mite *Galendromus helveolus* (Chant) (Acari: Phytoseiidae) has been reported to feed on eggs, larvae and nymphs of *B. californicus* [@Chen2006] and studies of *Amblyseius largoensis* (Muma) (Acari: Phytoseiidae) demonstrated its ability to suppress populations of *B. yothersi* on citrus plants [@Argolo2020]. In addition, the entomopathogenic fungus *Metarhizium anisopliae* var. *acridum* was found to be pathogenic to *B. phoenicis* [@Magalhaes2005]. In our own surveys, we have found mites which succumbed to fungal infections (`r citef("fig_3")`); however the fungal species was not determined. -->


### Conclusion
OFV is the type member of the genus *Dichorhavirus* (family *Rhabdoviridae*, order *Mononegavirales*) contain a bipartite, single-stranded and negative sense RNA genome. The dichorhavirus that infects citrus in Hawaii, Mexico, Colombia, and South Africa are identical to the OFV in gene order, content, and the genome sequence. According to the International committee on Taxonomy of Viruses (ICTV) classification, OFV consist of two orchid strains (OFV-Orc1 and OFV-Orc2) and two citrus strains (OFV-Cit1 and OFV-Cit2). Both the orchid strains of OFV infects citrus but none of the citrus strain have been reported from any orchid species. Detecting OFV in Florida represents a concern for horticulturists who grow orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagaceae species which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including cultivating *Vanilla* in southern Florida [@Chambers2019] and the famous Ghost Orchid, [*Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe]. The first time leprosis has been observed in the Florida in 1860’s and eradicated in the mid-1960s. In fact, @Kitajima2011 found that the Citrus Leprosis virus (CiLV) which previously affected Florida citrus was a nuclear type of citrus leprosis, which are closely related to OFV strains [@Roy2013]. Association of a distant relative of OFV which was named as Citrus leprosis dichorhavirus-N0 (CiLV-N0) was confirmed in relation to the leprosis disease outbreak in Florida [@Hartung2015]. The recent detection of OFV-Orc1 in South Africa [@Cook2019] in *C. sinensis* (Navel and Valencia orange) and OFV-Orc2 in Hawaii [@Velarde2021] in *C. reticulata* (mandarin) and *C. jambhiri* (rough lemon) highlights the threat of different strains of OFV; which will be a definite concern to the multi-billion dollar citrus industry. Lastly, some OFV isolates are known to be involved with citrus leprosis disease in Mexico [@Roy2015], which may be a cause for concern for the citrus industry. *B. californicus* and *B. yothersi* are both known vectors of Dichorhaviruses (OFV) and Cileviruses (Citrus Leprosis) [@Knorr1968; @Kondo2003; @Beltran-Beltran2020] and *B. obovatus* is a suspected vector as well [@Childers2003]. All three mite species/complexes are present in Florida [@Childers2003; @Akyazi2017]<!-- (`r citef("fig_4")`) -->. Therefore, it is critical to identify the vector of OFVs in Florida and monitor its spread to determine the risk this virus represents for the native plants, agriculture and the ornamental/landscaping industries of Florida and the surrounding regions.

### Acknowledgements
We would like to give a special thanks to the Tallahassee Museum for their patience, cooperation, and support with collecting plant samples. We also want to thank Drs. Sam Bolton, FDACS and Aline Tassi, Univ. of Sao Paulo, Brazil for checking the mites we have sent for species validation. We are especially indebted to the late Dr. Gary Bauchan for permitting us to use these beautiful Cryo-SEM images, he will be greatly missed. This research was partly funded by the USDA National Institute of Food and Agriculture, Hatch project FLA-NFC-005607. Mention of trade names or commercial products in this publication is solely for the purpose of providing specific information and does not imply recommendation or endorsement by the USDA; USDA is an equal opportunity provider and employer.

<!-- ### References -->
<!-- <div id="refs"></div> -->

<!-- ### Table -->

<!-- |Scientific Name|Common Names|Symptoms Observed| -->
<!-- |:-----------|:--------------|:----------------| -->
<!-- |*Liriope muscari* Bailey| Lilyturf, Orchardgrass, Monkeygrass| Ringspots, Chlorotic Flecking, Necrotic Lesions| -->
<!-- |*Liriope gigantea*\* Hume| Giant Lilyturf| Ringspots, Chlorotic Flecking, Necrotic Lesions| -->
<!-- |*Ophiopogon japonicus* Ker Gawl.| Dwarf Lilyturf, Mondo Grass, Snake's beard| Ringspots, Chlorotic Flecking, Necrotic Lesions| -->
<!-- |*Ophiopogon intermedius*\*\* Don| Aztec Grass, 'Argenteomarginatus' | Ringspots, Chlorotic Flecking, Necrotic Lesions| -->
<!-- |*Aspidistra elatior* Blume| Cast Iron Plant, Bar-room Plant | Chlorosis, Necrotic Lesions| -->
<!-- `r tab_nums(name = "tab_1", caption = "List of plants with symptoms of Orchid fleck dichorhavirus found in northern Florida. * *L. gigantea* have been traditionally classified as seperate from *L. muscari* by @Broussard2007 and @Fantz2015, although this distinction has been challenged by @Wang2014 and @Masiero2020. * * *O. intermedius* is sometimes misclassified as *Liriope muscari* ‘Variegated Evergreen Giant’ [@Fantz2008b, @Fantz2009] or ‘Grandiflora White’ [@Fantz2009].")` -->


### Figure captions
`r fig_nums(name = "fig_1", caption = "Variety of symptoms expressed by plants infected with Orchid fleck dichorhavirus: (a) ringspot symptoms on *Liriope gigantea* (b) chlorotic flecking on *Aspidistra elatior*.")`

<!-- `r fig_nums(name = "fig_2", caption = "LT-SEM images of *Brevipalpus californicus* sensu lato displaying various characters used for identification [@Baker1987; @Beard2012] (a) Dorsum  (b) Lateral view (c) Venter (d) Close up of distal end of leg 2, with arrows indicating paired solenidia, characteristic of the genus *Brevipalpus* (e) Enlargement of the microplates of the mite cerotegument (f) Dorsal view of the distal portion of mite abdomen (g) Dorsal view of the mite rostrum (h) Ventral view of mite rostrum, observe 3 distal setae. LT-SEM images provided by Dr. Gary Bauchan, USDA-ARS 2021")` -->

<!-- `r fig_nums(name = "fig_3", caption = "LT-SEM images of unidentified fungus infecting *Brevipalpus californicus* sensu lato: (a) Infested *B. californicus* adult, dorsal (b) Detail of fungal sporangia. LT-SEM images provided by Dr. Gary Bauchan, USDA-ARS 2021")` -->

<!-- `r fig_nums(name = "fig_4", caption = "Florida is home to other common pest species of *Brevipalpus* which are potential vectors of *Orchid fleck dichorhavirus*: (a) *B. yothersi*, dorsal (b) *B. yothersi*, lateral (c) *B. obovatus*, dorsal. SEM images provided by Dr. Ron Ochoa, USDA-ARS 2021")` -->

### Figures
```{r, fig_1, warning=FALSE}
grid::grid.raster(tiff::readTIFF( "images/fig_1.tif"))
```

<!-- ```{r, fig_2, warning=FALSE} -->
<!-- grid::grid.raster(tiff::readTIFF( "images/b_cali.tif")) -->
<!-- ``` -->

<!-- ```{r, fig_3, warning=FALSE} -->
<!-- grid::grid.raster(tiff::readTIFF("images/fungus_mite.tif")) -->
<!-- ``` -->

<!-- ```{r, fig_4, warning=FALSE} -->
<!-- grid::grid.raster(tiff::readTIFF( "images/other_mites.tif")) -->
<!-- ``` -->
